% Copyright 2007 by Mark Wibrow
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.


\section{Customizing the Mathematical Engine}
\label{pgfmath-reimplement}

Perhaps you have a desire for some function that \pgfname\ does not provide.
Perhaps you are not happy with the accuracy or efficiency of some of the
algorithms that are implemented in \pgfname. In these cases you will want to
add a function to the parser or replace the current implementations of the
algorithms with your own code.

The mathematical engine was designed with such customization in mind. It is
possible to add new functions, or modify the code for existing functions. Note,
however, that whilst adding new operators is possible, it can be a rather
tricky business and is only recommended for adventurous users.

To add a new function to the math engine the following command can be used:

\begin{command}{\pgfmathdeclarefunction\opt{|*|}\marg{name}\marg{number of arguments}\marg{code}}
    This will set up the parser to recognize a function called \meta{name}. The
    name of the function can consist of, uppercase or lowercase letters,
    numbers or the underscore |_|. In line with many programming languages, a
    function name cannot begin with a number or contain any spaces. The
    function may not have been declared earlier, unless the optional star (|*|)
    is provided, which forces an ``overwriting'' of the function by the new
    function. Note that you \emph{should never change the arity of standard
    functions} and you should normally use |\pgfmathredeclarefunction|,
    which does not allow you to do anything wrong here.

    The \meta{number of arguments} can be any positive integer, zero, or the
    value |...|, which indicates a variable number of arguments. \pgfname{}
    treats constants, such as |pi| and |e|, as functions with zero arguments.
    Functions with more than nine arguments or with a variable number of
    arguments are a ``bit special'' and are discussed below.

    The effect of \meta{code} should be to set the macro |\pgfmathresult| to
    the correct value (namely to the result of the computation without units).
    Furthermore, the function should have no other side effects, that is, it
    should not change any global values. As an example, consider the creation
    of a new function |double|, which takes one argument, and returns the value
    of that argument times two.
    %
\begin{codeexample}[]
\makeatletter
\pgfmathdeclarefunction{double}{1}{
  \begingroup
    \pgf@x=#1pt\relax
    \multiply\pgf@x by2\relax
    \pgfmathreturn\pgf@x
  \endgroup
}
\makeatother
\pgfmathparse{double(44.3)}\pgfmathresult
\end{codeexample}

    The macro |\pgfmathreturn|\meta{tokens} must be
    directly followed by an |\endgroup| and will save the result of the
    computation, by defining |\pgfmathresult| as the expansion of
    \meta{tokens} (without units) outside the group, so \meta{tokens}
    must be something that can be assigned to a dimension register.

    Alternatively, the |\pgfmathsmuggle|\meta{macro} can be used. This must
    also be directly followed by an |\endgroup| and will simply ``smuggle'' the
    definition of \meta{macro} outside the \TeX-group.

    By performing computations within a \TeX-group, \pgfname{} registers such
    as |\pgf@x|, |\pgf@y| and |\c@pgf@counta|, |\c@pgfcountb|, and so forth,
    can be used at will.

    Beyond setting up the parser, this command also defines two macros which
    provide access to the function independently of the parser:
    %
    \begin{itemize}
        \item |\pgfmath|\meta{function name}

            This macro will provide a ``public'' interface for the function
            \meta{function name} allowing the function to be called
            independently of the parser. All arguments passed to this macro are
            evaluated using |\pgfmathparse| and then passed on to the following
            macro:
        \item |\pgfmath|\meta{function name}|@|

            This macro is the ``private'' implementation of the function's
            algorithm (but note that, for speed, the parser calls this macro
            rather than the ``public'' one). Arguments passed to this macro are
            expected to be numbers without units. It is defined using
            \meta{code}, but need not be self-contained.
    \end{itemize}

    For functions that are declared with less than ten arguments, the public
    macro is defined in the same way as normal \TeX{} macros using, for
    example, |\def\pgfmathNoArgs{|\meta{code}|}| for a function with no
    arguments, or |\def\pgfmathThreeArgs#1#2#3{|\meta{code}|}| for a function
    with three arguments. The private macro is defined in the same way, and
    each argument can therefore be accessed in \meta{code} using |#1|, |#2| and
    so on.

    For functions with more than nine arguments, or functions with a variable
    number of arguments, these macros are only defined as taking \emph{one}
    argument. The public macro expects its arguments to be comma separated, for
    example, |\pgfmathVariableArgs{1.1,3.5,-1.5,2.6}|. Each argument is parsed
    and passed on to the private macro as follows:
    |\pgfmathVariableArgs@{{1.1}{3.5}{-1.5}{2.6}}|. This means that some
    ``extra work'' will be required to access each argument (although it is a
    fairly simple task).

    Note that there are two exceptions to this arrangement: the public versions
    of the |min| and |max| functions still take two arguments for compatibility
    with older versions, but each of these arguments can take several comma
    separated values.
\end{command}

To redefine a function use the following command:

\begin{command}{\pgfmathredeclarefunction\marg{function name}\marg{algorithm code}}
    This command redefines the |\pgfmath|\meta{function name}|@| macro with the
    new \meta{algorithm code}. See the description of the
    |\pgfmathdeclarefunction| for details. You cannot change the number of
    arguments for an existing function.
    %
\begin{codeexample}[]
\makeatletter
\pgfmathdeclarefunction{foo}{1}{
  \begingroup
    \pgf@x=#1pt\relax
    \multiply\pgf@x by2\relax
    \pgfmathreturn\pgf@x
  \endgroup
}
\pgfmathparse{foo(42)}\pgfmathresult
\pgfmathredeclarefunction{foo}{
  \begingroup
    \pgf@x=#1pt\relax
    \multiply\pgf@x by3\relax
    \pgfmathreturn\pgf@x
  \endgroup
}
\pgfmathparse{foo(42)}\pgfmathresult
\makeatother
\end{codeexample}
    %
\end{command}

    \pgfname{} uses the last known definition of a function within the
    prevailing scope, so it is possible for a function to be redefined locally.
    You should also remember that any |.sty| or |.tex| file containing any
    re-implementations should be loaded after |pgfmath|.

    In addition to the above commands, the following key is provided to quickly
    create simple ad hoc functions which can greatly improve the readability of
    code, and is particularly useful in \tikzname{}:

\begin{key}{/pgf/declare function=\meta{function definitions}}
    This key allows simple functions to be created locally. Its use is perhaps
    best illustrated by an example:
    %
\begin{codeexample}[]
\begin{tikzpicture}
  \draw [help lines] (0,0) grid (3,2);
  \draw [blue, thick, x=0.0085cm, y=1cm,
    declare function={
      sines(\t,\a,\b)=1 + 0.5*(sin(\t)+sin(\t*\a)+sin(\t*\b));
    }]
    plot [domain=0:360, samples=144, smooth] (\x,{sines(\x,3,5)});
\end{tikzpicture}
\end{codeexample}

    Each definition in \meta{function definitions} takes the form
    \meta{name}|(|\meta{arguments}|)=|\meta{definition}|;| (note the semicolon
    at the end, this is very important). If multiple functions are being
    defined, the semicolon is used to separate them (\emph{not} a comma). The
    function \meta{name} can be any name that is not already a function name in
    the current scope. The list of \meta{arguments} are commands such as |\x|,
    or |\y| (it is not possible to declare functions that take variable numbers
    of arguments using this key). If the function takes no arguments, then the
    parentheses need not be used. The \meta{definition} should be an expression
    that can be parsed by the mathematical engine and should use the commands
    specified in \meta{arguments}.

    When specifying multiple functions, functions that appear later on in
    \meta{function definitions} can refer to earlier functions:
    %
\begin{codeexample}[pre={\pgfmathsetseed{1}}]
\begin{tikzpicture}[
  declare function={
    excitation(\t,\w) = sin(\t*\w);
    noise             = rnd - 0.5;
    source(\t)        = excitation(\t,20) + noise;
    filter(\t)        = 1 - abs(sin(mod(\t, 90)));
    speech(\t)        = 1 + source(\t)*filter(\t);
  }
]
  \draw [help lines] (0,0) grid (3,2);
  \draw [blue, thick, x=0.0085cm, y=1cm] (0,1) --
    plot [domain=0:360, samples=144, smooth] (\x,{speech(\x)});
\end{tikzpicture}
\end{codeexample}
    %
\end{key}
