% Copyright 2006 by Till Tantau
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.


\section{Spy Library: Magnifying Parts of Pictures}
\label{section-library-spy}

\begin{tikzlibrary}{spy}
  The package defines options for creating pictures in which some part
  of the picture is repeated in another area in a magnified way (as if
  you were looking through a spyglass, hence the name).
\end{tikzlibrary}


\subsection{Magnifying a Part of a Picture}

The idea behind the |spy| library is to make is easy to create high-density
pictures in which some important parts are repeated somewhere, but
magnified as if you were looking through a spyglass:

\begin{codeexample}[]
\begin{tikzpicture}
  [spy using outlines={circle, magnification=4, size=2cm, connect spies}]

  \draw [help lines] (0,0) grid (3,2);

  \draw [decoration=Koch curve type 1]
    decorate { decorate{ decorate{ decorate{ (0,0) -- (2,0) }}}};

  \spy [red] on (1.6,0.3)
             in node [left] at (3.5,-1.25);

  \spy [blue, size=1cm] on (1,1)
              in node [right] at (0,-1.25);
\end{tikzpicture}
\end{codeexample}

\begin{codeexample}[]
\begin{tikzpicture}[spy using overlays={size=12mm}]
  \draw [decoration=Koch snowflake]
    decorate { decorate{ decorate{ decorate{ (0,0) -- (2,0) }}}};

  \spy [green,magnification=3] on (0.6,0.1) in node at (-0.3,-1);
  \spy [blue,magnification=5]  on (1,0.5)   in node at (1,-1);
  \spy [red,magnification=10]  on (1.6,0.1) in node at (2.3,-1);
\end{tikzpicture}
\end{codeexample}


Note that this magnification uses what is called a \emph{canvas
  transformation} in this manual: Everything is magnified, including
line width and text.

In order for ``spying'' to work, the picture obviously has to be drawn
several times: Once at its normal size and then again for each
``magnifying glass.'' Several keys and commands work in concert to
make this possible:
\begin{itemize}
\item You need to make \tikzname\ aware of the fact that a picture (or
  just a scope) is to be magnified. This is done by adding the special
  key |spy scope| to a |{scope}| or |{tikzpicture}| (which is also
  just a scope). Some special keys like |spy using outlines|
  implicitly set the |spy scope|.

\item Inside this scope you may then use the command |\spy|, which is
  only available inside such scopes (so there is no danger of your
  inadvertently using this command outside such a scope). This command
  has a special syntax and will (at some point) create two nodes: One
  node that shows the magnified picture (called the \emph{spy-in
    node}) and another node showing which part of the original picture
  is magnified (called the \emph{spy-on} node). The spy-in node is,
  indeed, a normal node, so it can have any shape or border that you
  like and you can apply all of \tikzname's advanced features to
  it. The only difference compared to a normal node is that instead of
  some ``text'' it contains a magnified version of the picture,
  clipped to the size of the node.

  The |\spy| command does not create the nodes immediately. Rather,
  the creation of these nodes is postponed till the end of the
  |spy scope| in which the |\spy| command is used. This is necessary
  since in order to repeat the whole scope inside the node containing
  the magnified version, the whole picture needs to be available when
  this node is created.
\end{itemize}

A basic question any library for ``magnifying things'' has to address
is how you specify which part of the picture is to be
magnified (the spy-on node) and where this magnified part is to be
shown (the spy-in node). There are two possible ways:
\begin{enumerate}
\item You specify the size and position of the spy-on node. Then the
  size of the spy-in node is determined by the size of the spy-on node
  and the magnification factor -- you can still decide where the
  spy-in node should be placed, but not its size.
\item Alternatively, you specify the size and position of the spy-in
  node. Then, similarly to the first case, the size of the spy-on node
  is determined implicitly and you can only decide where the
  spy-on node should be placed, but not its size.
\end{enumerate}

The |spy| library uses the second method: You specify the size and
position of the spy-in nodes, the sizes of the spy-on nodes are then
computed automatically.



\subsection{Spy Scopes}

\begin{key}{/tikz/spy scope=\meta{options} (default \normalfont empty)}
  This option may be used with a |{scope}| or any environment that
  creates such a scope internally (like |{tikzpicture}|). It has the
  following effects:
  \begin{itemize}
  \item It resets a number of graphic state parameters, including the
    color, line style, and other. This is necessary for technical
    reasons.
  \item It tells \tikzname\ that the content of the scope should be saved
    internally in a special box.
  \item It defines the command |\spy| so that it can be used inside
    the scope.
  \item At the end of the scope, the nodes belonging to the |\spy|
    commands used inside the scope are created.
  \item The \meta{options} are saved in an internal style. Each time
    |\spy| is used, these \meta{options} will be used.
  \item Three keys are defined that provide useful shortcuts:
    \begin{key}{/tikz/size=\meta{dimension}}
      Inside a |spy scope|, |size| this is a shortcut for |minimum size|.
    \end{key}
    \begin{key}{/tikz/height=\meta{dimension}}
      Inside a |spy scope|, |height| this is a shortcut for |minimum height|.
    \end{key}
    \begin{key}{/tikz/width=\meta{dimension}}
      Inside a |spy scope|, |width| this is a shortcut for |minimum width|.
    \end{key}
  \end{itemize}
  It is permissible to nest |spy scopes|. In this case, all |\spy|
  commands inside the inner |spy scope| only have an effect on
  material inside the scope, whereas |\spy| commands outside the inner
  |spy scope| but inside the outer |spy scope| allow you to ``spy on
  the spy.''

\begin{codeexample}[]
\begin{tikzpicture}
  [spy using outlines={rectangle, red, magnification=5,
                       size=1.5cm, connect spies}]

  \begin{scope}
    [spy using outlines={circle, blue,
                         magnification=3, size=1.5cm, connect spies}]
    \draw [help lines] (0,0) grid (3,2);

    \draw [decoration=Koch curve type 1]
      decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

    \spy on (1.6,0.3) in node (zoom) [left] at (3.5,-1.25);
  \end{scope}

  \spy on (zoom.north west) in node [right] at (0,-1.25);
\end{tikzpicture}
\end{codeexample}

\end{key}



\subsection{The Spy Command}

\begin{command}{\spy \opt{\oarg{options}} |on| \meta{coordinate}
    |in node| \meta{node options}|;|}
  This command can only be used inside a |spy scope|. Let us start with the syntax:
  \begin{itemize}
  \item The |\spy| command is not a special case of |\path|. Rather,
    it has a small parser of its own.
  \item Following the optional \meta{options}, you must write |on|,
    followed by a coordinate. This coordinate will be the center of
    the area that is to be magnified.
  \item Following the \meta{coordinate}, you must write |in node|
    followed by some \meta{node options}. The syntax for these options is the same
    as for a normal |node| path command, such as |[left]| or
    |(foo) [red] at (bar)|. \emph{However},  \meta{node options} are
    \emph{not} followed by a curly brace. Rather, the \meta{node
      options} must directly be followed by a semicolon.
  \end{itemize}
  The effect of this command is the following: The \meta{options},
  \meta{coordinate}, and \meta{node options} are stored internally
  till the end of the current
  |spy scope|. This means that, in particular, you can reference any node
  inside the |spy scope|, even if it is not yet defined when the
  |\spy| command is given. At the end of the current |spy scope|, two
  nodes are created, called the \emph{spy-in node} and the
  \emph{spy-on node}.
  \begin{itemize}
  \item The \emph{spy-in node} is the node that contains a magnified
    part of the picture (the node \emph{in} which we see on what we
    spy). This node is, indeed, a normal \tikzname\
    node, so you can use all standard options to style this node. In
    particular, you can specify a shape or a border color or a drop
    shadow or whatever. The only thing that is special about this node
    is that instead of containing some normal text, its ``text'' is
    the magnified picture.

    To be precise, the picture of the |spy scope| is scaled by a
    certain factor, specified by the |lens| or |magnification| options
    discussed below, and the shifted in such a way that the
    \meta{coordinate} lies at the center of the spy-on node.
  \item The \emph{spy-on node} is a node that is centered on the
    \meta{coordinate} and whose size reflects exactly the area shown
    inside the spy-in node (the node containing \emph{on} what we
    spy).
  \end{itemize}

  Let us now go over what happens in detail when the two nodes are
  created:
  \begin{enumerate}
  \item A scope is started. Two sets of options are used with this
    scope: First, the options passed to the enclosing |spy scope| and
    then the \meta{options} (which will, thus, overrule the options of
    the |spy scope|).
  \item Then, the spy-on node is created. However, we will first
    discuss the spy-in node.
  \item The spy-in node is created after the spy-on node (and, hence,
    will cover the spy-on node in case they overlap). When this node is
    created, the \meta{node options} are used in addition to the
    effect caused by the \meta{options} and the options of the
    |{spy scope}|. Additionally, the following style is used:
    \begin{stylekey}{/tikz/every spy in node}
      This style is used with every spy in node.
    \end{stylekey}
    The position of the node (the |at| option) is set to the
    \meta{coordinate} by default, so that it will cover the
    to-be-magnified area. You can change this by providing the |at|
    option yourself:
\begin{codeexample}[]
\begin{tikzpicture}
  [spy using outlines={circle, magnification=3, size=1cm}]

  \draw [decoration=Koch curve type 1]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [red]  on (1.6,0.3) in node;
  \spy [blue] on (1,1)     in node at (1,-1);
\end{tikzpicture}
\end{codeexample}
    No ``text'' can be specified for the node. Rather, the ``text''
    shown inside this node is the picture of the current |spy scope|,
    but canvas-transformed according to the following key:
    \begin{key}{/tikz/lens=\meta{options}}
      The \meta{options} should contain transformation commands like
      |scale| or |rotate|. These transformations are applied to the
      picture when it is shown inside the spy-on node.
    \end{key}
    Since the most common transformation is undoubtedly a simple
    scaling, there is a special style for this:
    \begin{key}{/tikz/magnification=\meta{number}}
      This has the same effect as saying
      |lens={scale=|\meta{number}|}|.
    \end{key}
    Now, usually the size of a node is determined in such a way that
    it ``fits'' around the text of the node. For a spy-on node this is
    not a good approach since the ``text'' of this node would contain
    ``the whole picture.'' Because of this, \tikzname\ acts
    as if the ``text'' of the node has zero size. You must then use
    keys like |minimum size| to cause the node to have a certain
    size. Note that the key |size| is an abbreviation for
    |minimum size| inside a spy scope.

    You can name the spy on node in the usual ways. Additionally, the
    node is (also) always named |tikzspyinnode|. Following the spy
    scope, you can use this node like any other node:
\begin{codeexample}[]
\begin{tikzpicture}
  \begin{scope}
    [spy using outlines={circle, magnification=3, size=2cm, connect spies}]

    \draw [decoration=Koch curve type 1]
      decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

    \spy [red] on (1.6,0.3) in node (a) [left] at (3.5,-1.25);

    \spy [blue, size=1cm] on (1,1) in node (b) [right] at (0,-1.25);
  \end{scope}
  \draw [ultra thick, green!50!black] (b) -- (a.north west);
\end{tikzpicture}
\end{codeexample}

  \item Once both nodes have been created, the current value of the
    following key is used to connect them:
    \begin{key}{/tikz/spy connection path=\meta{code} (initially
        \normalfont empty)}
      The \meta{code} is executed after the spy-on and spy-in nodes
      have just been created. Inside this \meta{code}, the two nodes
      can be accessed as |tikzspyinnode| and  |tikzspyonnode|.
      For example, the key |connect spies| sets this command to
\begin{codeexample}[code only]
\draw[thin] (tikzspyonnode) -- (tikzspyinnode);
\end{codeexample}
    \end{key}
  \end{enumerate}
  Returning to the creation of the spy-in node: This node is centered on
  \meta{coordinate} (more precisely, its anchor is set to |center| and
  the |at| option is set to \meta{coordinate}). Its size and shape are
  initially determined in the same way as the size and shape of the
  spy-on node (unless, of course, you explicitly provide a different
  shape for, say, the spy-on node locally, which is not really a good
  idea). Then, additionally, the \emph{inverted} transformation done
  by the |lens| option is applied, resulting in a node whose size and
  shape exactly corresponds to the area in the picture that is shown
  in the spy-on node.
\begin{codeexample}[]
\begin{tikzpicture}
  [spy using outlines={lens={scale=3,rotate=20}, size=2cm, connect spies}]

  \draw [decoration=Koch curve type 1]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [red] on (1.6,0.3) in node at (2.5,-1.25);
\end{tikzpicture}
\end{codeexample}

  Like for the spy-in node, a style can be used to format the spy-on
  node:
  \begin{stylekey}{/tikz/every spy on node}
    This style is used with every spy on node.
  \end{stylekey}
  The spy-on node is named |tikzspyonnode| (but, as always, this node
  is only available after the spy scope). If you have multiple
  spy-on nodes and you would like to access all of them, you need to
  use the |name| key inside the |every spy on node| style.

  The |inner sep| and |outer sep| of both spy-in and spy-on nodes are
  set to |0pt|.
\end{command}



\subsection{Predefined Spy Styles}

There are some predefined styles that make using the |spy| library
easier. The following two styles can be used instead of |spy scope|,
they pass their \meta{options} directly to |spy scope|. They
additionally setup the graphic styles to be used for the spy-in nodes
and the spy-on nodes in some special way.

\begin{key}{/tikz/spy using outlines=\meta{options} (default
    \normalfont empty)}
  This key creates a |spy scope| in which the spy-in node is drawn,
  but not filled, using a thick line; and the spy-on node is drawn,
  but not filled, using a very thin line.

\begin{codeexample}[]
\begin{tikzpicture}
  [spy using outlines={circle, magnification=3, size=1cm, connect spies}]

  \draw [decoration=Koch curve type 1]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [red] on (1.6,0.3) in node at (3,1);
\end{tikzpicture}
\end{codeexample}
\end{key}

\begin{key}{/tikz/spy using overlays=\meta{options} (default
    \normalfont empty)}
  This key creates a |spy scope| in which both the spy-in and spy-on
  nodes are filled, but with the fill opacity set to 20\%.

\begin{codeexample}[]
\begin{tikzpicture}
  [spy using overlays={circle, magnification=3, size=1cm, connect spies}]

  \draw [decoration=Koch curve type 1]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [green] on (1.6,0.3) in node at (3,1);
\end{tikzpicture}
\end{codeexample}
\end{key}

The following style is useful for connecting the spy-in and the spy-on
nodes:

\begin{key}{/tikz/connect spies}
  Causes the spy-in and the spy-on nodes to be connected by a thin
  line.

\begin{codeexample}[]
\begin{tikzpicture}
  [spy using overlays={circle, magnification=3, size=1cm}]

  \draw [decoration=Koch curve type 2]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [green] on (1.6,0.1) in node at (3,1);
  \spy [red,connect spies] on (0.5,0.4) in node at (1,1.5);
\end{tikzpicture}
\end{codeexample}
\end{key}


\subsection{Examples}

Usually, the spy-in node and the spy-on node should have the same
shape. However, you might also wish to use the |circle| shape for the
spy-on node and the |magnifying glass| shape for the spy-in node:

\begin{codeexample}[]
\tikzset{spy using mag glass/.style={
    spy scope={
      every spy on node/.style={
        circle,
        fill, fill opacity=0.2, text opacity=1},
      every spy in node/.style={
        magnifying glass, circular drop shadow,
        fill=white, draw, ultra thick, cap=round},
      #1
    }}}
\begin{tikzpicture}[spy using mag glass={magnification=3, size=1cm}]
  \draw [decoration=Koch curve type 2]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy [green!50!black] on (1.6,0.1) in node at (2.5,-0.5);
\end{tikzpicture}
\end{codeexample}

With the magnifying glass, you can also put it ``on top'' of the
picture itself:

\begin{codeexample}[]
\begin{tikzpicture}
  [spy scope={magnification=4, size=1cm},
   every spy in node/.style={
     magnifying glass, circular drop shadow,
     fill=white, draw, ultra thick, cap=round}]

  \draw [decoration=Koch curve type 2]
    decorate{ decorate{ decorate{ (0,0) -- (2,0) }}};

  \spy on (1.6,0.1) in node;
\end{tikzpicture}
\end{codeexample}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "pgfmanual-pdftex-version"
%%% End:
