% Copyright 2018 by Christian Feuersaenger
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.


\section{Adding libraries to \pgfname: temporary registers}
\label{section-internal-registers}

This section is intended for those who like to write libraries to extend
\pgfname. Of course, this requires a good deal of knowledge about
\TeX-programming and the structure of the \pgfname\ basic layer. Besides, one
will encounter the need of temporary variables and, especially, temporary \TeX\
registers. This section describes how to use a set of pre-allocated temporary
registers of the basic layer without needing to allocate more of them.

A part of these internals are already mentioned in
section~\ref{section-internal-pointcmds}, but the basic layer provides more
temporaries than |\pgf@x| and |\pgf@y|.

\begin{internallist}[dimen register]{\pgf@x,\pgf@y}
    These registers are used to process point coordinates in the basic layer of
    \pgfname, see section~\ref{section-internal-pointcmds}. After a
    |\pgfpoint|$\dotsc$ command, they contain the final $x$ and $y$ coordinate,
    respectively.

    The values of |\pgf@x| and |\pgf@y| are set \emph{globally} in contrast to
    other available \pgfname\ registers. You should never assume anything about
    their value unless the context defines them explicitly.

    Please prefer the |\pgf@xa|, |\pgf@xb|, $\dotsc$ registers for temporary
    dimen registers unless you are writing point coordinate commands.
\end{internallist}

\begin{internallist}[dimen register]{
    \pgf@xa,
    \pgf@xb,
    \pgf@xc,
    \pgf@ya,
    \pgf@yb,
    \pgf@yc%
}
    Temporary registers for \TeX\ dimensions which can be modified freely. Just
    make sure changes occur only within \TeX\ groups.

    \paragraph{Attention:}
    %
    \pgfname\ uses these registers to perform path operations. For reasons of
    efficiency, path commands do not always guard them. As a consequence, the
    code
    %
\begin{codeexample}[code only]
\pgfpointadd{\pgfpoint{\pgf@xa}{\pgf@ya}}{\pgfpoint{\pgf@xb}{\pgf@yb}}
\end{codeexample}
    %
    \noindent may fail: Inside |\pgfpointadd|, the |\pgf@xa| and friend
    registers might be modified. In particular, it might happen that |\pgf@xb|
    is changed before |\pgfpoint{\pgf@xb}{\pgf@yb}| is evaluated. The right
    thing to do would be to first expand everything using |\edef| and process
    the values afterwards, resulting in unnecessary expensive operations. Of
    course, one can avoid this by simply looking into the source code of
    |\pgfpointadd| to see which registers are used.
\end{internallist}

\begin{internallist}[dimen register]{\pgfutil@tempdima,\pgfutil@tempdimb}
    Further multi-purpose temporary dimen registers. For \LaTeX, these
    registers are already allocated as |\@tempdima| and |\@tempdimb| and are
    simply |\let| to the |\pgfutil@|$\dotsc$ names.
\end{internallist}

\begin{internallist}[count register]{
    \c@pgf@counta,
    \c@pgf@countb,
    \c@pgf@countc,
    \c@pgf@countd%
}
    These multiple-purpose count registers are used throughout \pgfname\ to
    perform integer computations. Feel free to use them as well, just make sure
    changes are scoped by local \TeX\ groups.
\end{internallist}

\begin{internallist}[openout handle]{\w@pgf@writea}
    An |\openout| handle which is used to generate complete output files within
    locally scoped parts of \pgfname\ (for example, to interact with
    |gnuplot|). You should always use |\immediate| in front of output
    operations involving |\w@pgf@writea| and you should always close the file
    before returning from your code.
    %
\begin{codeexample}[code only]
\immediate\openout\w@pgf@writea=myfile.dat
\immediate\write\w@pgf@writea{...}%
\immediate\write\w@pgf@writea{...}%
\immediate\closeout\w@pgf@writea%
\end{codeexample}
    %
\end{internallist}

\begin{internallist}[openin handle]{\r@pgf@reada}
    An |\openin| handle which is used to read files within locally scoped parts
    of \pgfname, for example to check if a file exists or to read data files.
    You should always use |\immediate| in front of output operations involving
    |\r@pgf@writea| and you should always close the file before returning from
    your code.
    %
\begin{codeexample}[code only]
\immediate\openin\r@pgf@reada=myfile.dat
% do something with \macro
\ifeof\r@pgf@reada
    % end of file or it doesn't exist
\else
    % loop or whatever
    \immediate\read\r@pgf@reada to\macro
    ...
\fi
\immediate\closein\r@pgf@reada
\end{codeexample}
    %
\end{internallist}

\begin{internallist}[box]{\pgfutil@tempboxa}
    A box for temporary use inside of local \TeX\ scopes. For \LaTeX, this box
    is the same as the already pre-allocated |\@tempboxa|.
\end{internallist}
