% Copyright 2006 by Till Tantau
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.



\section{Background Library}

\label{section-tikz-backgrounds}

\begin{tikzlibrary}{backgrounds}
  This library defines ``backgrounds'' for pictures. This does not
  refer to background pictures, but rather to frames drawn around and
  behind pictures. For example, this package allows you to just add
  the |framed| option to a picture to get a rectangular box around
  your picture or |gridded| to put a grid behind your picture.
\end{tikzlibrary}


The first use of this library is to make the following key available: 
\begin{key}{/tikz/on background layer}
  This key can be used with a |{scope}|. It will cause everything
  inside the scope to be typeset on a background layer. Note that the
  scope should not be ``deeply nested'' inside the picture since
  changes to the graphic state (like the color or the transformation
  matrix) ``do not survive a layer switch.'' For more details on
  layers see Section~\ref{section-layers}.

\begin{codeexample}[]
\begin{tikzpicture}
  % On main layer:
  \fill[blue] (0,0) circle (1cm);
  
  \begin{scope}[on background layer]
    \fill[yellow] (-1,-1) rectangle (1,1);
  \end{scope}
  
  \begin{scope}[on background layer]
    \fill[black] (-.8,-.8) rectangle (.8,.8);
  \end{scope}

  % On main layer again:
  \fill[blue!50] (-.5,-1) rectangle (.5,1);
\end{tikzpicture}
\end{codeexample}
\end{key}

When this package is loaded, the following styles become available:
\begin{stylekey}{/tikz/show background rectangle}
  This style causes a rectangle to be drawn behind your graphic. This
  style option must be given to the |{tikzpicture}| environment or to
  the |\tikz| command.
\begin{codeexample}[]
\begin{tikzpicture}[show background rectangle]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  The size of the background rectangle is determined as follows:
  We start with the bounding box of the picture. Then, a certain
  separator distance is added on the sides. This distance can be
  different for the $x$- and $y$-directions and can be set using the
  following options:
  \begin{key}{/tikz/inner frame xsep=\meta{dimension} (initially 1ex)}
    Sets the additional horizontal separator distance for the
    background rectangle.    
  \end{key}
  \begin{key}{/tikz/inner frame ysep=\meta{dimension} (initially 1ex)}
    Same for the vertical separator distance.
  \end{key}  
  \begin{key}{/tikz/inner frame sep=\meta{dimension}}
    Sets the horizontal and vertical separator distances
    simultaneously. 
  \end{key}
  The following two styles make setting the inner separator a bit
  easier to remember:
  \begin{stylekey}{/tikz/tight background}
    Sets the inner frame separator to 0pt. The background rectangle
    will have the size of the bounding box. 
  \end{stylekey}
  \begin{stylekey}{/tikz/loose background}
    Sets the inner frame separator to 2ex.
  \end{stylekey}  

  You can influence how the background rectangle is rendered by setting
  the following style:
  \begin{stylekey}{/tikz/background rectangle (initially draw)}
    This style dictates how the background rectangle is drawn or
    filled. The default setting causes the path of the background
    rectangle to be drawn in the usual 
    way. Setting this style to, say, |fill=blue!20| causes a light
    blue background to be added to the picture. You can also use more
    fancy settings as shown in the following example:
\begin{codeexample}[]
\begin{tikzpicture}
  [background rectangle/.style=
     {double,ultra thick,draw=red,top color=blue,rounded corners},  
   show background rectangle]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
    Naturally, no one in their right mind would use the above, but
    here is a nice background: 
\begin{codeexample}[]
\begin{tikzpicture}
  [background rectangle/.style=
     {draw=blue!50,fill=blue!20,rounded corners=1ex},
   show background rectangle]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  \end{stylekey}
\end{stylekey}

\begin{stylekey}{/tikz/framed}
  This is a shorthand for |show background rectangle|.
\end{stylekey}

\begin{stylekey}{/tikz/show background grid}
  This style behaves similarly to the |show background rectangle|
  style, but it will not use a rectangle path, but a grid. The lower
  left and upper right corner of the grid is computed in the same way
  as for the background rectangle:
\begin{codeexample}[]
\begin{tikzpicture}[show background grid]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  You can influence the background grid by setting
  the following style:
  \begin{stylekey}{/tikz/background grid (initially draw,help lines)}
    This style dictates how the background grid path is drawn. 
\begin{codeexample}[]
\begin{tikzpicture}
  [background grid/.style={thick,draw=red,step=.5cm},
   show background grid]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  \end{stylekey}
  This option can be combined with the |framed| option (use the
  |framed| option first):
\begin{codeexample}[]
\tikzset{background grid/.style={thick,draw=red,step=.5cm},
         background rectangle/.style={rounded corners,fill=yellow}}
\begin{tikzpicture}[framed,gridded]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
\end{stylekey}

\begin{stylekey}{/tikz/gridded}
  This is a shorthand for |show background grid|.
\end{stylekey}

\begin{stylekey}{/tikz/show background top}
  This style causes a single line to be drawn at the top of the
  background rectangle. Normally, the line coincides exactly with the
  top line of the background rectangle:
\begin{codeexample}[]
\begin{tikzpicture}[
    background rectangle/.style={fill=yellow},
    framed,show background top]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  The following option allows you to lengthen (or shorten) the line:
  \begin{key}{/tikz/outer frame xsep=\meta{dimension} (initially 0pt)}
    The \meta{dimension} is added at the left and right side of the
    line. 
\begin{codeexample}[]
\begin{tikzpicture}
  [background rectangle/.style={fill=yellow},
   framed,
   show background top,
   outer frame xsep=1ex]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  \end{key}
  \begin{key}{/tikz/outer frame ysep=\meta{dimension} (initially 0pt)}
    This option does not apply to the top line, but to the left and
    right lines, see below.
  \end{key}
  \begin{key}{/tikz/outer frame sep=\meta{dimension}}
    Sets both the $x$- and $y$-separation.
  \end{key}
\begin{codeexample}[]
\begin{tikzpicture}
  [background rectangle={fill=blue!20},
   outer frame sep=1ex,%
   show background top,%
   show background bottom,%
   show background left,%
   show background right]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  You can influence how the line is drawn grid by setting
  the following style:
  \begin{stylekey}{/tikz/background top (initially draw)}
\begin{codeexample}[]
\tikzset{background rectangle/.style={fill=blue!20},
         background top/.style={draw=blue!50,line width=1ex}}
\begin{tikzpicture}[framed,show background top]
  \draw (0,0) ellipse (10mm and 5mm);
\end{tikzpicture}
\end{codeexample}
  \end{stylekey}
\end{stylekey}

\begin{stylekey}{/tikz/show background bottom}
  Works like the style for the top line.
\end{stylekey}
\begin{stylekey}{/tikz/show background left}
  Works similarly.
\end{stylekey}
\begin{stylekey}{/tikz/show background right}
  Works similarly.
\end{stylekey}



%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "pgfmanual-pdftex-version"
%%% End: 
