% Copyright 2011 by Renée Ahrens, Olof Frahm, Jens Kluttig, Matthias Schulz, Stephan Schuster
% Copyright 2011 by Till Tantau
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.

\section{Graph Drawing Algorithms for Trees}


\begin{tikzlibrary}{graphdrawing.trees}
  Load this package when you wish to use graph drawing algorithms for
  trees. You should load the |graphdrawing| library first.
\end{tikzlibrary}

\ifluatex\relax\else{LuaTeX is required for setting this manual section.}\expandafter\endinput\fi


\begin{key}{/graph drawing/tree=\meta{options}}
  \keyalias{tikz}\keyalias{tikz/graphs}
  Similar to the |>| option, this ``generic'' name for a tree drawing
  algorithm is not hardwired to any specific algorithm. Rather, users
  can select an algorithm somewhere at the beginning of their program
  and then just write |\graph[tree]| to draw a tree.

  The \meta{options} will be forwarded to the currently selected
  algorithm.
\begin{codeexample}[]
\tikz \graph [tree] { a -> {b,c} };    
\end{codeexample}
  
  To change the algorithm, change the following key:
  \begin{key}{/graph drawing/tree/default algorithm=\meta{algorithm}}
    Set this key to the tree drawing algorithm of your choice. The
    default is |AhrensFKSS2011 tree|, but this will change.
  \end{key}
\end{key}


\begin{gdalgorithm}{AhrensFKSS2011 tree}
  This algorithm is due to Ren\'ee Ahrens, Olof Frahm, Jens Kluttig,
  Matthias Schulz, Stephan Schuster, who have implemented the graph
  drawing engine.
  
  This algorithm requires that the graph is a tree (but this may
  change in the future). By default, the root of the tree is
  positioned horizontally centered at the top. Child nodes are
  positioned beside each other according to their level. Here is an
  example: 

\begin{codeexample}[]
\tikz[AhrensFKSS2011 tree] \graph { a -> {b -> {c,d}, e}};
\end{codeexample}

  As you can see, the text nodes aren't quite aligned, so the common fix
  is to use the |text depth| and |text height| keys to force the text
  nodes to a specific size.

\begin{codeexample}[]
\tikz[AhrensFKSS2011 tree, text depth=.2em, text height=.8em]
  \graph { a -> {b -> {c,d}, e}};
\end{codeexample}

  \medskip
  \noindent\textbf{Parameters.} 
  The keys affecting the algorithm are the following common graph
  drawing parameters:

  \begin{key}{/graph drawing/root}
    \keyalias{tikz}\keyalias{tikz/graphs}
    This is a node parameter. At most one node should have this key
    set. If no node has it set, the first node in the graph will be
    used. 
  \end{key}

  \begin{key}{/graph drawing/level distance=\meta{leveldistance} (default 1cm)}
    \keyalias{tikz}\keyalias{tikz/graphs}
    Determines the vertical space between the nodes on different levels:
\begin{codeexample}[]
\tikz [AhrensFKSS2011 tree, level distance=1cm]
  \graph { 1 -> {2 , 3}};
\tikz [AhrensFKSS2011 tree, level distance=2cm]
  \graph { 1 -> {2 , 3}};
\end{codeexample}
  \end{key}

  \begin{key}{/graph drawing/sibling distance=\meta{siblingdistance} (default 1cm)}
    \keyalias{tikz}\keyalias{tikz/graphs}
    This determines the horizontal space between the nodes. 
\begin{codeexample}[]
\tikz [AhrensFKSS2011 tree, sibling distance=1cm]
  \graph { 1 ->{2 , 3}};
\tikz [AhrensFKSS2011 tree, sibling distance=2cm]
  \graph { 1 ->{2 , 3}};
\end{codeexample}
\end{key}

  \medskip
  \textbf{How does this algorithm work?}
  The tree algorithm works recursively. During the recursion one step is
  performed for each subgraph of the tree.  

  The process builds a kind of a box structure of the given graph. This
  means a leaf of a tree returns itself as a box. Its parent returns
  itself and its children in a bigger box etc. as shown in the following
  figure. 

\begin{quote}
\begin{tikzpicture}[
    level 0/.style={draw=black!50,very thick},
    level 1/.style={draw=orange!50,very thick},
    level 2/.style={draw=blue!50,very thick},
    level 3/.style={draw=green!50,very thick}]

    \node[level 1] (1) {1}
      child {node[level 2] (3) {3}
        child {node[level 3] (4) {4}
          child{node[level 3] (6) {6}}
          child{node[level 3] (7) {7}}
        }
        child {node[level 2] (5) {5}}
      }
      child {node[level 1] (2) {2}};

    \begin{pgfonlayer}{background}
        \node [level 0, fit=(1) (6) (2)] {};
        \node [level 1, fit=(3) (5) (6) (7)] {};
        \node [level 2,fit=(4) (6) (7)] {};
    \end{pgfonlayer}
\end{tikzpicture}
\end{quote}

  In each step the current boxes can be compared by their size, sorted
  and positioned. In the figure above the boxes of one step are
  represented in the same color. 
  
  In the tree algorithm the boxes of each tree level are first sorted
  ascendingly by their size and then arranged as follows: The the
  biggest box is positioned in the middle. Then the following boxes are
  positioned alternately left and right. 
  
  After this arrangement the relative coordinates for the position of
  each box have to be computed. The \emph{y}-coordinate of a box (except
  for the root node of the step) are determined by the maximum height of
  all boxes to guarantee a uniform layout of the tree. Nodes on the same
  level in the tree are positioned at the same height. The
  \emph{x}-coordinate of a box depends on the coordinates of its left
  neighbour box and an additional spacing (by default 10pt), which can
  be influenced by the |sibling distance| key. The \emph{y}-coordinate
  of the root node of each step is set to the maximum \emph{y}-value of
  the other boxes adding the same spacing meantioned above (by default
  10pt, influenced by |level distance|). Its \emph{x}-coordinate is
  determined by the width of the other boxes divided by 2. This means
  the root node is positioned in the middle above the other boxes. 
  
  Because each box knows its root node, it is possible to determine the
  absolute position of each box or node afterwards.  
  
  At the end of the step the current boxes are added to a result box and
  returned. 
\end{gdalgorithm}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "pgfmanual-pdftex-version"
%%% End: 
