% Copyright 2006 by Till Tantau
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.


\section{Layered Graphics}
\label{section-layers}

\subsection{Overview}

\pgfname\ provides a layering mechanism for composing graphics from multiple
layers. (This mechanism is not to be confused with the conceptual ``software
layers'' the \pgfname\ system is composed of.) Layers are often used in graphic
programs. The idea is that you can draw on the different layers in any order.
So you might start drawing something on the ``background'' layer, then
something on the ``foreground'' layer, then something on the ``middle'' layer,
and then something on the background layer once more, and so on. At the end, no
matter in which ordering you drew on the different layers, the layers are
``stacked on top of each other'' in a fixed ordering to produce the final
picture. Thus, anything drawn on the middle layer would come on top of
everything of the background layer.

Normally, you do not need to use different layers since you will have little
trouble ``ordering'' your graphic commands in such a way that layers are
superfluous. However, in certain situations you only ``know'' what you should
draw behind something else after the ``something else'' has been drawn.

For example, suppose you wish to draw a yellow background behind your picture.
The background should be as large as the bounding box of the picture, plus a
little border. If you know the size of the bounding box of the picture at its
beginning, this is easy to accomplish. However, in general this is not the case
and you need to create a ``background'' layer in addition to the standard
``main'' layer. Then, at the end of the picture, when the bounding box has been
established, you can add a rectangle of the appropriate size to the picture.


\subsection{Declaring Layers}

In \pgfname\ layers are referenced using names. The standard layer, which is a
bit special in certain ways, is called |main|. If nothing else is specified,
all graphic commands are added to the |main| layer. You can declare a new layer
using the following command:

\begin{command}{\pgfdeclarelayer\marg{name}}
    This command declares a layer named \meta{name} for later use. Mainly, this
    will set up some internal bookkeeping.
\end{command}

The next step toward using a layer is to tell \pgfname\ which layers will be
part of the actual picture and which will be their ordering. Thus, it is
possible to have more layers declared than are actually used.

\begin{command}{\pgfsetlayers\marg{layer list}}
    This command tells \pgfname\ which layers will be used in pictures. They
    are stacked on top of each other in the order given. The layer |main|
    should always be part of the list. Here is an example:
    %
\begin{codeexample}[code only]
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\end{codeexample}

    This command should be given either outside of any picture or ``directly
    inside'' of a picture. Here, the ``directly inside'' means that there
    should be no further level of \TeX\ grouping between |\pgfsetlayers| and
    the matching |\end{pgfpicture}| (no closing braces, no |\end{...}|). It
    will also work if |\pgfsetlayers| is provided before |\end{tikzpicture}|
    (with similar restrictions).
\end{command}


\subsection{Using Layers}

Once the layers of your picture have been declared, you can start to ``fill''
them. As said before, all graphics commands are normally added to the |main|
layer. Using the |{pgfonlayer}| environment, you can tell \pgfname\ that
certain commands should, instead, be added to the given layer.

\begin{environment}{{pgfonlayer}\marg{layer name}}
    The whole \meta{environment contents} is added to the layer with the name
    \meta{layer name}. This environment can be used anywhere inside a picture.
    Thus, even if it is used inside a |{pgfscope}| or a \TeX\ group, the
    contents will still be added to the ``whole'' picture. Using this
    environment multiple times inside the same picture will cause the
    \meta{environment contents} to accumulate.

    \emph{Note:} You can \emph{not} add anything to the |main| layer using this
    environment. The only way to add anything to the main layer is to give
    graphic commands outside all |{pgfonlayer}| environments.
    %
\begin{codeexample}[]
\pgfdeclarelayer{background layer}
\pgfdeclarelayer{foreground layer}
\pgfsetlayers{background layer,main,foreground layer}
\begin{tikzpicture}
  % On main layer:
  \fill[blue] (0,0) circle (1cm);

  \begin{pgfonlayer}{background layer}
    \fill[yellow] (-1,-1) rectangle (1,1);
  \end{pgfonlayer}

  \begin{pgfonlayer}{foreground layer}
    \node[white] {foreground};
  \end{pgfonlayer}

  \begin{pgfonlayer}{background layer}
    \fill[black] (-.8,-.8) rectangle (.8,.8);
  \end{pgfonlayer}

  % On main layer again:
  \fill[blue!50] (-.5,-1) rectangle (.5,1);
\end{tikzpicture}
\end{codeexample}
    %
\end{environment}

\begin{plainenvironment}{{pgfonlayer}\marg{layer name}}
    This is the plain \TeX\ version of the environment.
\end{plainenvironment}

\begin{contextenvironment}{{pgfonlayer}\marg{layer name}}
    This is the Con\TeX t version of the environment.
\end{contextenvironment}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "pgfmanual"
%%% End:
