% Copyright 2006 by Till Tantau
%
% This file may be distributed and/or modified
%
% 1. under the LaTeX Project Public License and/or
% 2. under the GNU Free Documentation License.
%
% See the file doc/generic/pgf/licenses/LICENSE for more details.


\section{Design of the System Layer}

\makeatletter

\subsection{Driver Files}
\label{section-pgfsys}

The \pgfname\ system layer mainly consists of a large number of commands
starting with |\pgfsys@|. These commands will be called \emph{system commands}
in the following. The higher layers ``interface'' with the system layer by
calling these commands. The higher layers should never use |\special| commands
directly or even check whether |\pdfoutput| is defined. Instead, all drawing
requests should be ``channeled'' through the system commands.

The system layer is loaded and set up by the following package:

\begin{package}{pgfsys}
    This file provides ``default implementations'' of all system commands, but
    most simply produce a warning that they are not implemented. The actual
    implementations of the system commands for a particular driver like, say,
    |pdftex| reside in files called |pgfsys-xxxx.sty|, where |xxxx| is the
    driver name. These will be called \emph{driver files} in the following.

    When |pgfsys.sty| is loaded, it will try to determine which driver is used
    by loading |pgf.cfg|. This file should set up the macro |\pgfsysdriver|
    appropriately. The |pgfsys.sty| will input the appropriate
    |pgfsys-|\meta{drivername}|.sty|.
\end{package}

\begin{command}{\pgfsysdriver}
    This macro should expand to the name of the driver to be used by |pgfsys|.
    The default from |pgf.cfg| is |pgfsys-\Gin@driver|. This is very likely to
    be correct if you are using \LaTeX. For plain \TeX, the macro will be set
    to |pgfsys-pdftex.def| if |pdftex| is used and to |pgfsys-dvips.def|
    otherwise.
\end{command}

\begin{filedescription}{pgf.cfg}
    This file should set up the command |\pgfsysdriver| correctly. If
    |\pgfsysdriver| is already set to some value, the driver normally should
    not change it. Otherwise, it should make a ``good guess'' at which driver
    will be appropriate.
\end{filedescription}

The currently supported backend drivers are discussed in
Section~\ref{section-drivers}.


\subsection{Common Definition Files}

Some drivers share many |\pgfsys@| commands. For the reason, files defining
these ``common'' commands are available. These files are \emph{not} usable
alone.

\begin{filedescription}{pgfsys-common-postscript}
    This file defines some |\pgfsys@| commands so that they produce appropriate
    PostScript code.
\end{filedescription}

\begin{filedescription}{pgfsys-common-pdf}
    This file defines some |\pgfsys@| commands so that they produce appropriate
    \textsc{pdf} code.
\end{filedescription}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "pgfmanual"
%%% End:
