% Copyright 2003 by Till Tantau <tantau@cs.tu-berlin.de>.
%
% This program can be redistributed and/or modified under the terms
% of the LaTeX Project Public License Distributed from CTAN
% archives in directory macros/latex/base/lppl.txt.

\section{Design of the System Layer}

\makeatletter


\subsection{Driver Files}
\label{section-pgfsys}

The \pgfname\ system layer consists of a large number of commands
starting with |\pgfsys@|. These commands will be called \emph{system
  commands} in the following. The higher layers ``interface'' with the
system layer by calling these commands. The higher layers should never
use |\special| commands directly or even check whether |\pdfoutput| is 
defined. Instead, all drawing requests should be ``channeled'' through
the system commands.

The system layer is loaded and setup by the following package:

\begin{package}{pgfsys}
  This file provides ``default implementations'' of all system
  commands, but most simply produce a warning that they are not
  implemented. The actual implementations of the system commands for a
  particular driver like, say, |pdftex| reside in files called
  |pgfsys-pdftex.sty|. These will be called \emph{driver files} in the
  following.

  When |pgfsys.sty| is loaded, it will try to determine which driver
  is used by loading |pgf.cfg|. This file should setup the macro
  |\pgfsysdriver| appropriately. The, |pgfsys.sty| will input the
  appropriate |pgfsys-|\meta{drivername}|.sty|. 
\end{package}

\begin{command}{\pgfsysdriver}
  This macro should expand to the name of the driver to be used by
  |pgfsys|. The default from |pgf.cfg| is |pgfsys-\Gin@driver|. This
  is very likely to be correct if you are using \LaTeX. For plain
  \TeX, the macro will be set to |pgfsys-pdftex.def| if |pdftex| is
  used und to |pgfsys-dvips.def| otherwise.
\end{command}

\begin{filedescription}{pgf.cfg}
  This file should setup the command |\pgfsysdriver| correctly. If
  |\pgfsysdriver| is already set to some value, the driver normally
  should not change it. Otherwise, it should make a ``good guess'' at
  which driver will be appropriate.
\end{filedescription}



\subsection{System Commands Shared Between Different Drivers}

Some definitions of system layer commands can be ``shared'' between
different drivers. For example, the literal text needed to stroke a
path in pdf is |S|, independently of the driver. For this reason,
the drivers for |pfdtex| and for |dvipdfm|, both of which produce
|.pdf| in the end, both include the file |pgfsys-common-pdf.def|,
which defines all common commands. Similarly, all PostScript based
drivers can used |pgfsys-common-postscript.def| for the ``standard''
postscript commands.


\subsection{Existing Driver Files}

With the current version of \pgfname, the following drivers are
implemented:

\subsection{Supported Drivers}

\begin{filedescription}{pgfsys-pdftex.def}
  This is a driver file for use with pdf\TeX, that is, with the
  |pdftex| or |pdflatex| command. It includes
  |pgfsys-common-pdf.def|. This driver has the most functionality. 
\end{filedescription}

\begin{filedescription}{pgfsys-dvipdfm.def}
  This is a driver file for use with (|la|)|tex| followed by |dvipdfm|. It
  includes |pgfsys-common-pdf.def|. This driver uses |graphicx| for the
  graphics inclusion and does not support masking. It does not
  support image inclusion in plain \TeX\ mode.
\end{filedescription}

\begin{filedescription}{pgfsys-dvips.def}
  This is a driver file for use with (|la|)|tet| followed by
  |dvips|. It includes |pgfsys-common-postscript.def|. This driver
  uses |graphicx| for the graphics inclusion and does not support
  masking. Shading is implemented, but the results will not be
  as good as with a driver producing |.pdf| as output. It does not
  support image inclusion in plain \TeX\ mode.
\end{filedescription}


\subsection{Common Definition Files}

Some drivers share many |\pgfsys@| commands. For the reason, files
defining these ``common'' commands are available. These files are
\emph{not} usable alone.

\begin{filedescription}{pgfsys-common-postscript}
  This file defines |\pgfsys@| commands so that they produce
  appropriate PostScript code.
\end{filedescription}

\begin{filedescription}{pgfsys-common-pdf}
  This file defines |\pgfsys@| commands so that they produce
  appropriate \textsc{pdf} code.
\end{filedescription}

